\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\emergencystretch=1.5em
\usepackage[hidelinks]{hyperref}

\newcommand{\ZZ}{\mathbb{Z}}
\theoremstyle{plain}
\newtheorem*{proposition}{Proposition}
\newtheorem*{lstheorem}{Theorem 2}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\title{Two remarks on Bedert's ``On unique sums in Abelian groups''}
\author{Mark Watson}
\date{21 July 2026}

\begin{document}
\maketitle

This note records two observations about Bedert's paper [B]. First, the upper bound
$m(p) = O((\log p)^2)$ of [B, Theorem 5] already follows, with a worse constant,
from a theorem of {\L}uczak and Schoen [LS], which is [B]'s own reference [9]
(Section 1). Second,
[B, Proposition 1] holds with its first binomial factor removed (Section 2); this
is the one result proved in full here, and it is machine-checked in Lean, apart
from its final, elementary Vandermonde step. Section 3 collects smaller observations from a close
reading of [B]'s proofs, about information that the packaged statements discard,
together with an elementary limiting construction and some computations. Everything else is context, and is [B]'s or
[LS]'s.

For $A \subseteq \ZZ/p\ZZ$ and $g \in \ZZ/p\ZZ$, write
\[
r_A(g) \;=\; |\{(a,a') \in A^2 : a + a' = g\}|,
\]
counting ordered pairs. $A$ has \emph{no unique sum} iff no $g$ satisfies
$1 \le r_A(g) \le 2$. Following [B], $m(p)$ is the size of a smallest
$A \subseteq \ZZ/p\ZZ$ with no unique sum; I take $|A| \ge 2$ throughout, which
excludes the vacuous $A = \emptyset$.

\section{$m(p) = O((\log p)^2)$ follows from [LS, Theorem 2] with $Q = 2$}

[LS] work with the same ordered count as $r_A$. They set
\[
\nu(t) \;=\; \nu_A(t) \;=\; |\{(a,b) \in A \times A : t = a+b\}|,
\qquad
\nu(A) \;=\; \min_{t \in A+A} \nu(t),
\]
with the convention (their p.~101) that ``if $a \neq b$, then we view $t = a+b$ and
$t = b+a$ as two different representations of $t$''. Their Theorem 2 (p.~109) states:

\begin{lstheorem}[{[LS], p.~109}]
``For every positive integer $Q < \log_2 p/(2\log_2(\gamma_p \log_2 p))$, where
$\gamma_p$ is the constant given in Straus' construction, there exists a set
$A \subseteq \ZZ/p\ZZ$ such that $|A| = (\gamma_p \log_2 p)^Q$ and
$\nu(A) \ge 2^Q$.''
\end{lstheorem}

\noindent They record $\gamma_p \le 2$, tending to $2/\log_2 3$.

Take $Q = 2$. The hypothesis reads
\[
\log_2 p \;>\; 4 \log_2(\gamma_p \log_2 p),
\]
which holds for $p \ge 2^{22}$. The resulting $A$ has
\[
|A| \;\le\; 4(\log_2 p)^2, \qquad \nu(A) \;\ge\; 4,
\]
the first bound using $\gamma_p \le 2$. Every $g$ with $r_A(g) \ge 1$ lies in $A + A$, so $r_A(g) = \nu_A(g) \ge 4 > 2$: no
$g$ has $r_A(g) \in \{1, 2\}$, and $A$ has no unique sum. Hence
\[
m(p) \;\le\; 4(\log_2 p)^2 \qquad (p \ge 2^{22}).
\]

This is the bound of [B, Theorem 5]. [LS] never state the $m(p)$ corollary, though
the step from $\nu \ge 4$ to ``no unique sum'' is immediate, and [B] (p.~3) records
the previous best upper bound as ``$m(p) \ll \sqrt{p}$ which came from a rather easy
construction''.

Bedert's construction is a constant-factor improvement over the one above, using
balanced sets in place of Straus's $S$. At $Q = 2$ the [LS] set is $S + x \cdot S$,
where [LS, Lemma 4] supplies an $x$ making $(s,s') \mapsto s + xs'$ injective, so
$|S + xS| = |S|^2$: the same two-dimensional shape as Bedert's $B \times B$, though
[LS, Lemma 4] gives injectivity only, not the higher-order condition a Freiman
isomorphism requires.

\section{A redundant factor in [B, Proposition 1]}

Let $Z = (z_1, \dots, z_n)$ be a finite indexed family in an abelian group, let
\[
\Sigma(Z) \;=\; \Bigl\{\, \textstyle\sum_{i \in I} z_i \;:\; I \subseteq [n] \,\Bigr\}
\]
be its set of subset sums, and let $d = \dim(Z)$ be the largest size of an index set
whose subset sums are pairwise distinct (equivalently, indexing a dissociated
subfamily). The trivial bound $|\Sigma(Z)| \le 2^n$ always holds, with equality
when $Z$ itself is dissociated, and [B, Proposition 1] shows this is essentially the
only way $\Sigma(Z)$ gets large: a big additive span forces a big dissociated
subfamily. Quantitatively, it states
\[
|\Sigma(Z)| \;\le\; \binom{n}{d}\binom{n+d}{d}.
\]
The observation here is that the first factor can be dropped.

\begin{proposition}
\begin{equation}
|\Sigma(Z)| \;\le\; \#\{I \subseteq [n] : I \text{ dissociated}\}
\;\le\; \sum_{j=0}^{d} \binom{n}{j} \;\le\; \binom{n+d}{d}.
\label{eq:chain}
\end{equation}
\end{proposition}

\begin{proof}
The idea is to pick one canonical index set for each attainable sum, and to count
the representatives by shattering. Order index sets by their characteristic
$0$--$1$ strings, lexicographically. For
each $y \in \Sigma(Z)$ let $R_y$ be the lex-least $I$ with $\sum_{i \in I} z_i = y$,
and set
\[
F \;=\; \{R_y : y \in \Sigma(Z)\},
\]
so that $|F| = |\Sigma(Z)|$.

Every set $S$ shattered by $F$ is dissociated. Otherwise there are distinct
$P, Q \subseteq S$ with equal sums; cancelling $P \cap Q$ we may take them disjoint,
say $P < Q$ lexicographically. Shattering gives $R \in F$ with $R \cap S = Q$; then
\[
R' \;=\; (R \setminus Q) \cup P
\]
has the same sum as $R$ and is lex-smaller, contradicting the minimality of $R$.

By Pajor's strengthening [P] of the Sauer--Shelah lemma [S, Sh], a finite family
shatters at
least as many sets as it has members, so $F$ shatters at least $|F|$ sets, all
dissociated. A dissociated index set has at most $d$ elements by definition of
$\dim$, and
\[
\sum_{j \le d} \binom{n}{j} \;\le\; \binom{n+d}{d}
\]
by Vandermonde.
\end{proof}

The argument uses only the indexed structure, so it covers multisets, the setting of
[B].

Bedert's own extremal example ([B] p.~6: $Z$ consists of $k$ copies of each of the
standard basis vectors $e_1, \dots, e_d$ of $\ZZ^d$, $n = kd$) does not obstruct the
improvement: there the first inequality of \eqref{eq:chain} is an equality, both
sides being $(k+1)^d$, since a dissociated index set picks at most one of the $k$
copies of each $e_i$.

Downstream, with $K = n/d$, [B, Corollary 1] becomes
\[
|\Sigma(Z)| \;\le\; \binom{n+d}{d} \;\le\; \bigl(e(K+1)\bigr)^d,
\]
of $K^d$ scale rather than the $K^{2d}$ of the stated $(4K)^{2d}$.

\begin{remark}
The first two inequalities of \eqref{eq:chain} are machine-checked in Lean, as
\texttt{BedertLab.card\_subsetSums\_le} in the repository accompanying this note,
with Mathlib's \texttt{Finset.card\_le\_card\_shatterer} as the classical input (the
Lean proof uses the colexicographic order; the argument is the same). The
final Vandermonde step is not formalized; \texttt{verify.py} checks it numerically.
\end{remark}

\section{Further observations}

These are secondary, none of them is asserted as new, and a remark after each one
states exactly what backs it.

The first three concern the lower-bound machinery of [B], so a word of context. That proof
runs an iteration (Propositions 5 and 6) in which a set $S_j$ of shifts is
repeatedly enlarged while $D + S_j$ covers more of $A$, for a fixed dissociated
$D \subseteq A$. Coverage grows by at least $|D|^2/(36|A|)$ per step, so the
iteration stops within $36K^2$ steps, where $K = |A|/\dim(A)$; and it can only stop
once $|S_j|$ has outgrown the constraint [B, (22)]. A lower bound on the number of
steps is therefore a lower bound on $K$. [B] writes $\omega_1$ for the resulting
bound: applying Proposition 5 (to a translate of $A$, as in the proof of Theorem 4)
gives $K \ge \omega_1(p)$, with
$\omega_1(p) \gg (\log\log\log p)^{1/2}$, and the conversion in the proof of his
Theorem 4 turns this into the $\omega(p)$ of the main result.

\subsection{The shift set in Proposition 6 grows polynomially}

[B, (23)] records only $|S'| \le \max(2|S|, |S|^3)$, so the iteration in Proposition
5 knows only $\log_2|S_j| \le 3^j$. Reading the proof, however, $S'$ always has one of two
forms:
\[
S' \;=\; S \cup (S+t) \ \text{ for a single } t,
\qquad\text{or}\qquad
S' \;=\; 2S - S \quad (\text{using } 0 \in S);
\]
these are the ``first case'' and ``final case'' of the proof, pp.~24--25.

Read as generalized arithmetic progressions, the two operations are ``append a side
of length $2$'' and ``map every side $L$ to $3L-2$'', and tracking the sides through
the iteration gives
\begin{equation}
\log_2 |S_j| \;\le\; 0.4\, j^2 + j,
\label{eq:growth}
\end{equation}
polynomial rather than exponential in $j$. With the same termination argument this
yields
\[
\omega_1(p) \;\gg\; (\log\log p)^{1/4}
\qquad\text{in place of}\qquad
(\log\log\log p)^{1/2}.
\]
In words: with $|S_j|$ growing only polynomially, far more steps are needed before
[B, (22)] fails, and every extra step raises the lower bound on $K$.

The tracked progression is never a hypothesis, only an observation about the sets
the proof already constructs, so no lemma in Proposition 6 is affected.

\begin{remark}
\texttt{verify.py} checks \eqref{eq:growth} for all $2^j$ schedules with $j \le 14$.
A Lean proof for arbitrary schedules, and a kernel-checked derivation of the
exponent from \eqref{eq:growth}, exist in my working development but are not
included in the repository; the passage from the two-shape reading of Proposition 6
to \eqref{eq:growth} is on paper only. The two-shape claim and its exhaustiveness
were read from the PDF.
\end{remark}

\subsection{The two branches of Proposition 6 have different increments}

[B, (24)] records an increment of $|D|^2/(36|A|)$ for both branches. Inspection of
the proof, however, gives
\[
\frac{|D|^2}{36|A|} \ \text{ for the translate branch},
\qquad
\frac{|D|^2}{12|A|} \ \text{ for the cube branch},
\]
three times larger, and (24) takes the minimum; nothing downstream restores the
difference, since Proposition 5 consumes only the packaged statement. With $c$
translate and $f$ cube steps the termination budget is therefore
\[
c + 3f \;\le\; 36K^2
\qquad\text{rather than}\qquad
c + f \;\le\; 36K^2,
\]
which divides the quadratic coefficient in the growth analysis by $3$ and multiplies
the final constant by exactly $3^{1/4} \approx 1.316$.

\begin{remark}
Read from the PDF's displayed formulas by two independent passes reaching identical
constants; not formalized.
\end{remark}

\subsection{What the $\omega_1$ improvement is worth}

The gain to $\omega_1$ is immediate and large; the gain to the published $\omega$ is
not. [B]'s conversion
\[
f(x) \;=\; \frac{x}{2(2 + \log_2 x)}
\]
penalizes a larger $\omega_1$ more, because its denominator grows with $x$. At
$\log_2\log_2 p = 2^{64}$, [B]'s $\omega_1 \approx 1.06$ loses a factor $4.2$ under
$f$, while $\omega_1 \approx 1.4 \times 10^4$ loses a factor $31.5$.

Comparing the published constant-free shapes,
\[
\frac{\sqrt{\log\log\log p}}{\log\log\log\log p}
\qquad\text{against}\qquad
\frac{(\log\log p)^{1/4}}{\log\log\log p},
\]
the crossover past which the improved $\omega$ exceeds [B]'s is around
$\log\log p \approx 5 \times 10^4$, that is $p \approx \exp(\exp(50000))$. So this is
an asymptotic improvement to $\omega$ and an immediate one only to $\omega_1$.

\begin{remark}
The numbers in this subsection are recomputed by \texttt{verify.py}.
\end{remark}

\subsection{A limit on the Section 2 improvement}

The Proposition of Section 2 removes one factor; the remaining $K^d$ scale cannot be
removed, even for sets. Take $d = 2h$ and
\[
Z \;=\; \{\, e_i + \chi_S \;:\; 1 \le i \le h,\ S \subseteq \{h+1,\dots,2h\} \,\}
\;\subset\; \ZZ^d,
\]
a set, not a multiset. Its subset sums record, in the first $h$ coordinates, how
many elements with each index were chosen, giving
\[
\log_2|\Sigma(Z)| \;\ge\; d^2/4.
\]
But any dissociated $D \subseteq \{0,1\}^d$ has $2^{|D|}$ distinct subset sums lying
in $\{0,\dots,|D|\}^d$, so
\[
2^{|D|} \;\le\; (|D|+1)^d,
\]
and $\dim(Z) = O(d\log d)$. The ratio diverges, so no bound of the form
$|\Sigma(Z)| \le 2^{C\dim(Z)}$ holds for sets.

\begin{remark}
Elementary; both halves are reproduced in \texttt{verify.py}.
\end{remark}

\subsection{Exact values of $m(p)$}

Computed by exact solver, certified optimal:
\[
\begin{array}{c|ccccccccccc}
p & 3 & 5 & 7 & 11 & 13 & 17 & 19 & 23 & 29 & 31 & 37\\\hline
m(p) & 3 & 4 & 5 & 7 & 7 & 8 & 9 & 10 & 11 & 11 & 12
\end{array}
\]
with $m(41), m(43) \le 13$. I did not find this sequence in the OEIS.

The solver's optimal witnesses for $p \le 23$ happen to be double-covering additive
bases, though at $p = 19$ and $23$ most extremal sets are not additive bases at all;
$m(p)$ tracks $2\sqrt{p}$ throughout this range. That is a warning rather than a
finding: the $2\sqrt{p}$ design regime governs until roughly $p \sim 10^3$ to
$10^4$, so these values cannot identify the asymptotic exponent, and fitting a
growth law to $p < 1000$ fits the wrong regime.

Separately, the minimum size of a balanced set is
$\lceil \log_2 p + 1\rceil + \{0,1\}$ throughout the computed range, so the
Bilu--Lev--Ruzsa rectification threshold $\log_2 p$ ([B, Lemma 1]) is essentially
sharp here.

\begin{remark}
The values of $m(p)$ for $p \le 19$, and the balanced-set minima on the same range,
were reproduced by independent exhaustive search, which \texttt{verify.py} repeats.
\end{remark}

\section*{Closing remarks}

Section 2 is a direct application of a classical theorem (Sauer [S], Shelah [Sh],
Pajor [P]), and while I could not locate it stated as a subset-sum bound, that is
the profile of a fact a specialist may regard as standard. I am not affiliated with
an institution, so MathSciNet and some publisher citation indices were unavailable
to me. I made substantial use of AI tools in this work; the quotations were checked
character for character against the sources, and the Proposition of Section 2 is
machine-checked in Lean, apart from the final Vandermonde step.

To check, in the repository
(\texttt{github.com/mkwatson/unique-sums-notes}): run
\texttt{lake exe cache get \&\& lake build}, then
\begin{center}
\texttt{\#print axioms BedertLab.card\_subsetSums\_le},
\end{center}
expecting \texttt{[propext, Classical.choice, Quot.sound]}; and run
\texttt{python3 verify.py} for the arithmetic and the small computations.

\begin{thebibliography}{Sh}

\bibitem[B]{bedert} B.~Bedert, \emph{On unique sums in Abelian groups},
Combinatorica \textbf{44}(2) (2024), 269--298; arXiv:2303.15134. Page references
are to the arXiv PDF.

\bibitem[LS]{ls} T.~{\L}uczak and T.~Schoen, \emph{On a problem of Konyagin},
Acta Arith.\ \textbf{134}(2) (2008), 101--109.

\bibitem[P]{pajor} A.~Pajor, \emph{Sous-espaces $l_1^n$ des espaces de Banach},
Travaux en Cours 16, Hermann, Paris, 1985.

\bibitem[S]{sauer} N.~Sauer, \emph{On the density of families of sets},
J.~Combin.\ Theory Ser.~A \textbf{13} (1972), 145--147.

\bibitem[Sh]{shelah} S.~Shelah, \emph{A combinatorial problem; stability and
order for models and theories in infinitary languages}, Pacific J.~Math.\
\textbf{41} (1972), 247--261.

\end{thebibliography}

\end{document}
